Detecting and quantifying parasite-induced host mortality from intensity data: method comparisons and limitations

### 2015-08-25 docs/supplementary_material.tex
Converting manuscript and supplementary to .doc

\noindent
\Large{\textbf{Supplementary Data}}

\normalsize

\section{Implementation of the Crofton Method}

@@

Second, choose a truncation value ($t_2$) such that $t_2 <
p_{max}$. Truncate the data such that $\text{data}_\text{truncated} <= t_2$ and repeat the above
iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
truncated value $t_3 < t_2$ and repeat the first two steps. Continue to truncate
the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
Because the method attempts to estimate three parameters, at least 3 classes are
needed for all 3 parameters to be identifiable \citep{Royce1990}.

@@

\section{Implementation of the Adjei Method}

@@

By taking the first and second derivatives of equation \ref{eq:logistic}, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.  This is in many ways analogous to the parasite pathogenicity parameter $\alpha$ given in classic macroparasite models, which specifies the slope of the linear relationship between between host death rate and parasite intensity \citep{AndersonandMay1978, Isham1995}.  The parameter $a$ is easily interpreted by holding $b$ constant and looking at how a one unit change in $a$ affects the log parasite intensity at which some percentage $p$ of hosts experience mortality.  Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation \ref{eq:logistic} gives.

\begin{align*}
    \log{\frac{p}{1 - p}} &= a_1 - b \log{x_1}\\
    - \log{\frac{p}{1 - p}} &= a_2 - b \log{x_2}\\
    0 &= a_1 - a_2 - b\log{x_1} + b\log{x_2}\\
    a_2 - a_1 &= b(\log{x_2} - \log{x_1})
\end{align*}
If $a_2 - a_1 = 1$, then the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.

To estimate the parameters in equation \ref{eq:logistic}, the Adjei Method first
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \mu_p, k_p)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation \ref{eq:logistic}.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

\section{Code and unit tests for estimating parasite-induced host mortality}

@@

\singlespacing
\bibliographystyle{int_j_para.bst}
\bibliography{parasite_host_mortality}

>>>>

\noindent
\textbf{Supplementary Data}

\textbf{1 Implementation of the Crofton Method}

@@

Second, choose a truncation value ($t_2$) such that $t_2 <
p_{max}$. Truncate the data such that $\text{data}_\text{truncated} \leq t_2$ and repeat the above
iterative procedure to calculate another set of parameters $(N_{p2}, \mu_{p2}, k_{p2})$
that minimizes the $\chi^2$-squared statistic on the truncated data. Choose a new
truncated value $t_3 < t_2$ and repeat the first two steps. Continue to truncate
the dataset until it only contains hosts with 0, 1, and 2 parasites (or 3 bins).
Because the method attempts to estimate three parameters, at least 3 classes are
needed for all 3 parameters to be identifiable \citep{Royce1990}.

@@

\textbf{2 Implementation of the Adjei Method}

@@

By taking the first and second derivatives of equation 1, one can easily find that the maximum rate of decline in host survival probability with increasing parasite intensity occurs at the $LD_{50}$ and has a value of $b / 4$.  This is in many ways analogous to the parasite pathogenicity parameter $\alpha$ given in classic macroparasite models, which specifies the slope of the linear relationship between between host death rate and parasite intensity \citep{AndersonandMay1978, Isham1995}.  The parameter $a$ is easily interpreted by holding $b$ constant and looking at how a one unit change in $a$ affects the log parasite intensity at which some percentage $p$ of hosts experience mortality.  Letting $a_1$ and $a_2$ be two different values of $a$ and $x_1$ and $x_2$ be two different parasite intensities, a bit of rearranging of equation 1 gives

$$\log{\frac{p}{1 - p}} = a_1 - b \log{x_1}$$
$$- \log{\frac{p}{1 - p}} = a_2 - b \log{x_2}$$
$$0 = a_1 - a_2 - b\log{x_1} + b\log{x_2}$$
$$a_2 - a_1 = b(\log{x_2} - \log{x_1})$$

If $a_2 - a_1 = 1$, then the change in log parasite intensity at which $p$ percentage of hosts survive is $1 / b$.

To estimate the parameters in equation 1, the Adjei Method first
calculates the expected number of hosts with a given parasite load $x$ by using
the equation $g(x ; \mu_p, k_p) * N_p$, where $g(x ; \mu_p, k_p)$ is the negative binomial pre-mortality distribution.  Second, the observed and predicted number of hosts
with $x$ parasites are paired as a single data point and the method then assumes that
this data point follows a binomial distribution with the total number of
``trials'' equal to the predicted number of hosts and the total number of
``successes'' equal to the observed number of hosts. In some cases, the
observed number of hosts is greater than the expected number of hosts and the
Adjei Method alters the data so that the observed is equal to the predicted
\citep{Adjei1986}.  After this questionable manipulation, the (observed, predicted) pairs are fit to a standard Generalized Linear Model \citep{McCullagh1989} with a binomial response variable and a logistic link function given by equation 1.  This model provides estimates for parameters $a$, $b$ and $LD_{50}$.

@@

\textbf{3 Code and unit tests for estimating parasite-induced host mortality}

@@

\textbf{References}

% \singlespacing
% \bibliographystyle{int_j_para.bst}
% \bibliography{parasite_host_mortality}
